---
title: "wrangling-and-cleaning.qmd"
format: html
editor: visual
---

## Load Libraries

```{r}
library(dplyr)
library(tidyr)
library(readr)

```

## Load Data

bit.ly/data-wrangle-lesson-data

```{r}

catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.30.1")
```

```{r}
colnames(catch_original)
```

```{r}
head(catch_original)
```

```{r}
summary(catch_original)
```

```{r}
unique(catch_original$Region)
```

```{r}
#View(catch_original)
```

```{r}
# %>% %>% %>% %>% ctl + shift + m %>% 
```

## Pipe Operator

## Select Columns

GOAL: remove 'All' and 'notersRedCode' columns

```{r}
catch_data <- catch_original %>% 
  select("Region","Year","Chinook","Sockeye","Coho","Pink","Chum")
```

## Quality Check

```{r}
dplyr::glimpse(catch_data)
```

## 

## Change column content

Goals: convert the Chinook values to numeric and overwrite

```{r}
catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))
```

## Find the row with the 'NA' value

Compare nested code versus piped code

```{r}
#which(is.na(catch_clean$Chinook))
catch_clean$Chinook %>% 
  is.na %>%  
  which()

```

Goal: change the value of the 'Chinook column to 1 if the value is equal to the letter I.

```{r}
catch_clean <- catch_data %>% 
  mutate(Chinook = if_else(condition = Chinook == "I",
                            true = "1",
                            false = Chinook))

catch_clean <- catch_clean %>%
  mutate(Chinook = as.numeric(Chinook))
```

```{r}
dplyr::glimpse(catch_clean)

```


Goal: transform dta from wide format to a long format


```{r}
catch_long <- catch_clean %>% 
  pivot_longer(
    cols = -c(Region, Year),
    names_to = "species",
          values_to = "catch"
          )

catch_long %>% 
  pivot_wider(names_from = species,
              values_from = catch)
  
```

## Renaming columns

Goal: rename the 'catch' comn to be 

```{r}
catch_long <- catch_long %>% 
  rename(catch_thousands = catch)


```


Goal: change units

```{r}
catch_long <- catch_long %>% 
  mutate(catch = catch_thousands * 1000)
```

Goal: remove catch_thousands

```{r}
catch_long <- catch_long %>% 
  select(-catch_thousands)
```

```{r}
head(catch_long)
```

## Summary Statistics

Goal: exercises

```{r}
#catch_long %>% filter(catch > 10000000)
catch_long %>% filter(Region == "SSE", species =="Chinook")
```

## Separate and Unite

```{r}
## create `cities_df`
cities_df <- data.frame(city = c("Juneau AK",
                                 "Sitka AK",
                                 "Anchorage AK"))


cit_st <- cities_df %>%  separate(city,c("city","state")," ")
cit_st

citst <- cit_st %>%  unite(city,state, sep = "-")
citst
```
```{r}
dates_df <- data.frame(
    year = c("1930",
             "1930",
             "1930"),
    month = c("12",
              "12",
              "12"),
    day = c("14",
            "15",
            "16")
)

dates_df %>%
    unite(date, year, month, day, sep = "-")
```

